Fail Faster

### 2025-03-20 paper/paper.tex
polishing s3

In Muli-stage programming (also known more simply as staging), programs execute
over the course of multiple stages, with each stage producing the code to be run
in the next. For the purposes of this paper, we will
primarily consider two stages, namely compile time and run time. Staged programs thus
execute twice: once at compile time, which produces more code, which is then
compiled and run at run time.

\subsubsection{Uses of Staging in eDSL Construction}
One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
\jwc{Cite racket macros for DSLs here:}
While embedded DSLs are powerful tools, they have well-known drawback: 
the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
This effect is known as ``abstraction overhead'' \cite{carrette05, moller20}: the layers of abstraction that 
make eDSLs so usable are precisely what prevents them from being fast.
Indeed, the root causes of many of the issues we discovered in Section~\ref{section:motiv} are
not unique to generator DSLs.
In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
This recipe has been used to great effect across domains to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
and query processing \cite{rhyme}.

Many languages have some degree of staging functionality, though support varies
from full native support (Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, \jwc{Racket??}) to compiler extensions (OCaml \cite{metaocaml,macocaml}, Java \cite{mint}).
For this paper, we have implemented staged PBT libraries both in OCaml --- using the MetaOCaml fork of the OCaml compiler \cn --- and Scala 3 --- using
\texttt{scala.quoted}. For presentation purposes, all staged code presented in the body of this paper is \camlname{} code written in OCaml:
\scalaname{} differs almost entirely in syntax alone.
For a further discussion of the potential of building staged generator DSLs in other languages, see Section~\ref{section:other-langs}.

\subsubsection{Staging in MetaOCaml}
MetaOCaml's staging functionality is exposed through a type \texttt{'a code}. A
value of type \texttt{t code} at compile time is a (potentially open) OCaml term
of type \texttt{t}.

Values of \texttt{code} type are introduced by \emph{quotes}, written
\texttt{.<$\ldots$>.}. Brackets delay execution of a program until run time. For
example, the program \texttt{.< 5 + 1 >.} has type \texttt{int code}.  Note that
this is not the same as \texttt{.< 6 >.}. Because brackets delay computation,
the code is not \emph{executed} until the next stage (run time).
Values of type \texttt{code} can be combined together using \emph{escape},
written \texttt{.\~{}(e)} (or just \texttt{.\~{}x}, when \texttt{x} is a variable).  Escape lets you take a value of type
\texttt{code}, and ``splice'' it directly into a quote.  For example, this
program \texttt{let x = .<1 * 5>. in .< .\~{}(x) + .\~{}(x) >.} evaluates to
\texttt{.<(1 * 5) + (1 * 5)>.}.  MetaOCaml ensures correct scoping and macro
hygiene, ensuring that variables are not shadowed when open terms are spliced.

The power of staging for optimization away abstraction overheads comes from
defining functions that accept and return \texttt{code} values.  A function
\texttt{f : 'a code -> 'b code} is a function that takes a program computing a
run-time \texttt{'a} and transforms it into a program computing a run-time
\texttt{'b}. In particular, because \texttt{f} itself runs at compile time, the
fact that the programmer called \texttt{f} does not matter at run time --- the abstraction
that \texttt{f} defines has been eliminated.
A code-transforming function \texttt{'a code -> 'b code} can also be converted \emph{code for a function} --- a value of type
\texttt{('a -> 'b) code} --- with the following program:
\begin{lstlisting}
let eta (f : 'a code -> 'b code) : ('a -> 'b) code = .<fun x -> .\~(f .<x>.)>.
\end{lstlisting}

This program is known as ``the trick'' in the partial evaluation and multi-stage programming literature.
\cn{} 
% USE THIS: https://arxiv.org/pdf/2309.08207
It returns a code for a function that takes an argument \texttt{x}, and then it splats in the result of calling \texttt{f} on just
the quoted \texttt{x}.

The trick is best illustrated by an example. The following program reduces to \texttt{.< fun x -> (1 + x) mod 2 == 0 >}.
\begin{lstlisting}
let is_even x = .< .~x mod 2 == 0 >. in
let succ x = .< 1 + .~x >. in
eta (succ . is_even)
\end{lstlisting}
By composing the two code-transforming functions together at compile time, and only then turning them into a run-time function,
the functions are fused together.
This is the basis of how staging is used to eliminate the abstraction of DSLs
(like a generator DSL). By writing DSL combinators as compile-time functions ---
and only calling \texttt{eta} at the end on the completed DSL program ---  we
can ensure that any overhead of using the combinators is eliminated by run time.

@@

To build a staged version of a generator DSL, we must rewrite the the generator
combinators so that they run during the compile time. Of course,
the compile time stage does not do all the work: the DSL must run at compile time,
producing inlined code that uses no combinators, which can then be compiled and run at run time
with different sizes and seeds.

Our job is thus to carefully bisect the DSL, determining which inputs to generator
combinators are known statically (and can be part of the compile time stage), and
which parts are only known at run time (and must be \texttt{code}).
In the staging literature, this task is known as ``binding-time analysis'', and it is more of an art than a science.

@@

The crux of our binding time analysis is that the \emph{only} parts of a generator that are not
known at compile time are the particular random seed and
size parameters that will be passed in at run time. Generators themselves are always known entirely at compile time.
This means our library's generator type \texttt{'a
Gen.t} should have the type \texttt{int code -> SR.t code -> 'a code}: a compile-time
function from dynamically-known size and seed to dynamically-determined result.

This type, along with basic staged monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
The constant generator \texttt{return} has the same code, but this time it 
runs at compile time. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it
returns the generator which always generates that value. Similarly,
\texttt{bind g k} sequences generators as usual by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one \footnote{Readers familiar with OCaml may notice that \texttt{return : 'a code -> 'a Gen.t} and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t}
do not have the correct types for a monad instance, preventing us from using \texttt{let\%bind} notation. We rectify this issue in Section~\ref{subsection:codecps} by importing some clever ideas from the staging literature.}.
Operationally, bind takes code for the size and seed, and returns code that (1) let-binds a variable \texttt{a} to spliced-in code that runs \texttt{g}, and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size\_c random\_c}
and \texttt{k .<a>. size\_c random\_c} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the randomness library.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}
from Section~\ref{subsection:msp}).
\iflater
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}
\fi

@@

  let to_bq (g : 'a code Gen.t) : ('a Bq.t) code =
  .<
    fun size random -> .~(g .<size>. .<random>.)
  >.
end
\end{lstlisting}
\caption{Basic Staged Generator Library}
\ref{fig:gen-staged-basic}
\end{figure}

@@

let int_pair : (int * int) Bq.t code = Gen.to_bq int_pair_staged
(* .< fun size random ->
        let x = SR.int random 0 100 in
        let y = SR.int random 0 x in
        (x,y)
    >.
*)
\end{lstlisting}
\caption{Pairs of Ints, Staged}
\ref{fig:running-staged}
\end{figure}

>>>>

In muli-stage programming (also known more simply as staging), programs execute
in multiple stages, with each stage producing the code to be run
in the next. For the purposes of this paper, we will
consider two stages: namely compile time and run time. In our context, staged programs
execute once at compile time, which produces more code, which is then
compiled and run at run time.

\paragraph{Uses of Staging in eDSL Construction}
One of the primary uses of staging is in embedded DSL construction \cite{sheard99, lms, trattdsl}.
\jwc{Cite racket macros for DSLs here:}
While embedded DSLs are powerful tools, they have well-known drawback: 
the functional abstractions used to build eDSLs prevent compilers from generating efficient machine code \cn.
This is abstraction overhead \cite{carrette05, moller20}: the layers of abstraction that 
make eDSLs so usable are precisely what prevents them from being fast.
Indeed, the root causes of many of the performance issues we discovered in Section~\ref{section:motiv} are
not unique to generator DSLs.
In light of this issue, staging is often used as a lightweight compiler for DSLs. The compile-time evaluation stage
transforms the DSL code, eliminating abstractions to produce code that the host language compiler can generate fast machine code for.
This recipe has been used to great effect to stage eDSLs for stream processing \cite{moller20,strymonas}, parser combinators \cite{sspc, staged-parsers, krishnaswami19, flap},
and query processing \cite{rhyme}.

Many languages have some degree of staging functionality, including
Scala 3 \cite{scalamacros}, Haskell \cite{templatehaskell}, \jwc{Racket??}, OCaml \cite{metaocaml,macocaml}, and Java \cite{mint}.
For this paper, we have implemented staged PBT libraries both in OCaml --- using the MetaOCaml fork of the OCaml compiler \cn --- and Scala 3 --- using
\texttt{scala.quoted}. For presentation purposes, all staged code presented in the body of this paper is \camlname{} code written in OCaml:
\scalaname{} is very similar.
For further discussion of the potential for building staged generator DSLs in other languages, see Section~\ref{section:other-langs}.

\paragraph{Staging in MetaOCaml}
MetaOCaml's staging functionality is exposed through the type \texttt{'a code}. A
value of type \texttt{t code} is an OCaml term
of type \texttt{t}.
Values of \texttt{code} type are introduced by \emph{quotes}, written
\texttt{.<$\ldots$>.}. Quotes delay execution of a program until run time. For
example, the program \texttt{.< 5 + 1 >.} has type \texttt{int code}.  Note that
this is not the same as \texttt{.< 6 >.}. Because brackets delay computation,
the code is not \emph{executed} until the next stage (run time).
Values of type \texttt{code} can be combined together using an \emph{escape},
written \texttt{.\~{}(e)} (or just \texttt{.\~{}x}, when \texttt{x} is a variable).  Escaping lets you take a value of type
\texttt{code}, and ``splice'' it directly into a quote.  For example, the
program \texttt{let x = .<1 * 5>. in .< .\~{}(x) + .\~{}(x) >.} evaluates to
\texttt{.<(1 * 5) + (1 * 5)>.}.  MetaOCaml ensures correct scoping and macro
hygiene, ensuring that variables are not shadowed when open terms are spliced.

The power of staging for optimizing away abstraction overheads comes from
defining functions that accept and return \texttt{code} values.  A function
\texttt{f : 'a code -> 'b code} is a function that takes a program computing a
run-time \texttt{'a} and transforms it into a program computing a run-time
\texttt{'b}. In particular, because \texttt{f} itself runs at compile time, the
fact that the programmer called \texttt{f} does not matter at run time --- the abstraction
that \texttt{f} defines has been eliminated.
A code-transforming function \texttt{'a code -> 'b code} can also be converted \emph{code for a function} --- a value of type
\texttt{('a -> 'b) code} --- with the following program:
\begin{lstlisting}
let eta (f : 'a code -> 'b code) : ('a -> 'b) code = .<fun x -> .\~(f .<x>.)>.
\end{lstlisting}

This program is known as ``the trick'' in the partial evaluation and multi-stage programming literature.
\cn{} 
% USE THIS: https://arxiv.org/pdf/2309.08207
It returns a code for a function that takes an argument \texttt{x} and splices in the result of calling \texttt{f} on just
the quoted \texttt{x}.

The trick is best illustrated by an example. The following program reduces to \texttt{.< fun x -> (1 + x) mod 2 == 0 >}.
\begin{lstlisting}
let is_even x = .< .~x mod 2 == 0 >. in
let succ x = .< 1 + .~x >. in
eta (fun x -> succ (is_even x))
\end{lstlisting}
By composing the two code-transforming functions together at compile time, and only then turning them into a run-time function,
the functions are fused together.
This is the basis of how staging is used to eliminate the abstraction of DSLs
(like a generator DSL). By writing DSL combinators as compile-time functions ---
and only calling \texttt{eta} at the end on the completed DSL program ---  we
can ensure that any overhead of using the combinators is eliminated by run time.
\jwc{BCP: You wrote a ``?'' next to this section. What's the issue?}

@@

To build a staged version of a generator DSL, we must rewrite the the generator
combinators to do as much as possible at compile time, 
producing inlined code that uses no combinators, which can be compiled and run
with different sizes and seeds.

Our job is thus to carefully bisect the DSL, determining which inputs to generator
combinators are known statically (and can be part of the compile time stage) and
which parts are only known at run time (and must be \texttt{code}).
In the staging literature, this task is known as ``binding-time analysis'' and it is more an art than a science.

@@

The crux of our binding time analysis is that the \emph{only} parts of a generator that are not
known at compile time are the particular random seed and
size parameters that will be passed in at run time. Generators themselves are always known entirely at compile time.
This leads us to define our library's generator type \texttt{'a
Gen.t} as \texttt{int code -> SR.t code -> 'a code}: a compile-time
function from dynamically-known size and seed to dynamically-determined result.

The basic staged monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
The constant generator \texttt{return} has the same code, but this time it 
runs at compile time. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it
returns the generator which always generates that value. Similarly,
\texttt{bind g k} sequences generators as usual by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one \footnote{Readers familiar with OCaml may notice that \texttt{return : 'a code -> 'a Gen.t} and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t}
do not have the correct types for a monad instance, preventing us from using \texttt{let\%bind} notation. We rectify this issue in Section~\ref{subsection:codecps} by importing some clever ideas from the staging literature.}.
Operationally, bind takes code for the size and seed, and returns code that (1) let-binds a variable \texttt{a} to spliced-in code that runs \texttt{g}, and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size\_c random\_c}
and \texttt{k .<a>. size\_c random\_c} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the randomness library.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}
from Section~\ref{subsection:msp}).
\iflater
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}
\fi

@@

  let to_bq (g : 'a code Gen.t) : ('a Bq.t) code =
  .<
    fun size random -> .~(g .<size>. .<random>.)
  >.
end
\end{lstlisting}
\caption{Basic Staged Generator Library}
\label{fig:gen-staged-basic}
\end{figure}

@@

let int_pair : (int * int) Bq.t code = Gen.to_bq int_pair_staged
(* .< fun size random ->
        let x = SR.int random 0 100 in
        let y = SR.int random 0 x in
        (x,y)
    >.
*)
\end{lstlisting}
\caption{Pairs of Ints, Staged}
\label{fig:running-staged}
\end{figure}
